\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-10-14 Paper/sachs2015plotROC.tex
compiled

\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. I reviewed a sample of
  ROC curve plots from the major oncology journals in order to assess
  current trends in usage and design elements. My review suggests that ROC
  curve plots are often ineffective as statistical charts and that poor
  design obscures the relevant information the chart is intended to
  display. I describe my new \proglang{R} package that was created to
  address the shortcomings of existing tools. The package has functions to
  create informative ROC curve plots, with sensible defaults, for use in
  print or as an interactive web-based plot. A web application was
  developed to reach a broader audience of scientists who do not use
  \proglang{R}.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves have
also been used for decades in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering two
possible types of errors: false positives, and false negatives. For a
continuous measurement that I denote as \(M\), convention dictates that
a test positive is defined as \(M\) exceeding some fixed threshold
\(c\): \(M > c\). In reference to the binary outcome that I denote as
\(D\), a good outcome of the test is when the test is positive for an
individual who truly has a disease: \(D = 1\). A bad outcome is when the
test is positive for an individual who does not have the disease
\(D = 0\).

@@

\[ TPF(c) = P\{ M > c | D = 1 \} \]

@@

\[ FPF(c) = P\{ M > c | D = 0 \} \]

Since the cutoff \(c\) is not fixed in advance, I can plot the TPF
against the FPF for all possible values of \(c\). This is exactly what
the ROC curve is, a plot of \(FPF(c)\) on the \(x\) axis and \(TPF(c)\)
along the \(y\) axis as \(c\) varies. A useless test that is not
informative at all in regards to the disease status has
\(TPF(c) = FPF(c)\) for all \(c\). The ROC plot of a useless test is
thus the diagonal line. A perfect test that is completely informative
about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for all
\(c\).

@@

\[ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i > c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, \]

@@

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
the interesting approach of encoding the cutoff values as a separate
color scale along the ROC curve itself. A legend for the color scale is
placed along the vertical axis on the right of the plotting region. Due
to the popularity of \pkg{ROCR}, my package supports the performance
objects as returned by \pkg{ROCR}. The \pkg{pROC} package \citep{pROC}
is mainly focused on estimating confidence intervals and regions for
restricted ranges of the ROC curve. The plotting methods therein use the
base \proglang{R} plotting functions to create nice displays of the
curves along with shaded confidence regions. My \pkg{plotROC} package
uses the \pkg{ggplot2} \citep{ggplot2} plotting library to create clear,
informative ROC plots, with interactive features for use on the web, and
sensible defaults for use in print.

\subsection{Motivation}\label{motivation}

@@

I start by creating an example data set. The marker I generate is
moderately accurate for predicting disease status. Next I use the
\code{calculate_roc} function to compute the empirical ROC curve. The
disease status need not be coded as 0/1, but if it is not,
\code{calculate_roc} assumes (with a warning) that the lowest value in
sort order signifies disease-free status. This returns a dataframe with
three columns: the cutoff values, the TPF and the FPF.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, size = 1, prob = .5)
R> M.ex <- rnorm(100, mean = D.ex)
R> roc.estimate <- calculate_roc(M.ex, D.ex)
R> head(roc.estimate)
\end{Sinput}
\begin{Soutput}
          c TPF       FPF
1 -2.182331   1 0.9811321
2 -1.674012   1 0.9622642
3 -1.568181   1 0.9433962
4 -1.550585   1 0.9245283
5 -1.539929   1 0.9056604
6 -1.190555   1 0.8867925
\end{Soutput}
\end{Schunk}

The \code{roc.estimate} object is passed to the \code{ggroc} function
with an optional label. This creates a \code{ggplot} object of the ROC
curve using the \pkg{ggplot2} package \citep{ggplot2}.

\begin{Schunk}
\begin{Sinput}
R> single.rocplot <- ggroc(roc.estimate, label = "Example")
\end{Sinput}
\end{Schunk}

The \code{single.rocplot} object can be used to create an interactive
plot and display it in the Rstudio viewer or default web browser by
passing it to the \code{plot_interactive_roc} function. Give the
function an optional path to an html file as an argument called
\code{file} to save the interactive plot as a complete web page. A
screen shot of an interactive plot is shown in Figure \ref{interact}.
Hovering over the display shows the cutoff value at the point nearest to
the cursor. Clicking makes the cutoff label stick until the next click,
and if confidence regions are available, clicks will also display those
as grey rectangles.

\begin{Schunk}
\begin{Sinput}
R> plot_interactive_roc(single.rocplot)
\end{Sinput}
\end{Schunk}

@@

An interactive ROC plot can be exported by using the
\code{export_interactive_roc} function, which returns a character string
containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
character string can be copy-pasted into an html document, or better
yet, incorporated directly into a dynamic document using \pkg{knitr}
\citep{knitr}. In a \pkg{knitr} document, it is necessary to use the
\code{cat} function on the results and use the chunk options
\code{results = 'asis'} and \code{fig.keep = 'none'} so that the
interactive plot is displayed correctly. For documents that contain
multiple interactive plots, it is necessary to assign each plot a unique
name using the \code{prefix} argument of \code{export_interactive_roc}.
This is necessary to ensure that the \proglang{JavaScript} code
manipulates the correct svg elements. For examples of interactive plots
and how to incorporate them into \pkg{knitr} documents, see the package
vignette (\code{vignette("examples", package = "plotROC")}) or the web
page https://sachsmc.github.io/plotROC/. The next code block shows an
example \pkg{knitr} chunk that can be used in an .Rmd document to
display an interactive plot.

\begin{Code}
```{r int-no, results = 'asis', fig.keep = 'none'}
cat(
  export_interactive_roc(single.rocplot, 
                        prefix = "a")
  )
```
\end{Code}

The same \code{ggroc} object that I called \code{single.rocplot} can be
used to generate an ROC plot suitable for use in print. It annotates the
cutoff values and is completely in black and white. A simple example
with the default options is shown in Figure \ref{first}.

\begin{Schunk}
\begin{Sinput}
R> plot_journal_roc(single.rocplot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
\end{figure}
\end{Schunk}

@@

If you have multiple tests of different types measured on the same
subjects, you can use the \code{calculate_multi_roc} function to compute
the empirical ROC curve for each test. It returns a list of data frames
with the estimates and cutoff values. Then the \code{multi_ggroc}
function creates the appropriate type of \code{ggplot} object.
Confidence regions are not supported for multiple curves at the time of
writing.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, 1, .5)
R> 
R> paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
+                        M2 = rnorm(100, mean = D.ex, sd = .4), 
+                        M3 = runif(100), D = D.ex)
R> 
R> estimate.list <- calculate_multi_roc(paired.data, 
+                                      c("M1", "M2", "M3"), "D")
\end{Sinput}
\end{Schunk}

Labels can be added easily with the \code{label} option of
\code{multi_ggroc}. The length of the label element should match the
number of plotted curves. The \code{multi_ggroc} object can be passed to
the \code{plot_journal_roc} and the \code{export_interactive_roc}
functions, as desired. The resulting plot is shown in Figure
\ref{multi}.

\begin{Schunk}
\begin{Sinput}
R> multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
R> plot_journal_roc(multi.rocplot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/multi2-1} \caption{Illustration of \pkg{plotROC} plot with multiple curves. \label{multi} }\label{fig:multi2}
\end{figure}
\end{Schunk}

Both \code{plot_journal_roc} and \code{export_interactive_roc} support a
number of options to customize the look of the plots. By default,
multiple curves are distinguished by different line types and direct
labels. For multiple ROC curves that are similar to one another, those
defaults can make it difficult to interpret the plots. Therefore, we
also support colors and legends. The x- and y-axis can be changed by
passing options to \code{ggroc} or \code{multi_ggroc}. The next code
block illustrates the available options.

\begin{Schunk}
\begin{Sinput}
R> colorplot <- multi_ggroc(estimate.list, 
+                          xlabel = "1 - Specificity", 
+                          ylabel = "Sensitivity")
R> cat(
+   export_interactive_roc(colorplot, lty = rep(1, 3), 
+                          color = c("black", "purple", "orange"), 
+                          legend = TRUE)
+   )
\end{Sinput}
\end{Schunk}

@@

\subsubsection{Click to view confidence
region}\label{click-to-view-confidence-region}

I use the \code{ci = TRUE} option in \code{calculcate_roc} and
\code{ggroc} to compute confidence regions for points on the ROC curve
using the \citet{clopper1934use} exact method. Briefly, exact confidence
intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
\citet{pepe2003statistical}, the cross-product of these intervals yields
a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
pair. The significance level can be specified using the \code{alpha}
option.

\begin{Schunk}
\begin{Sinput}
R> roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
+                         ci = TRUE, alpha = 0.05)
R> ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
\end{Sinput}
\end{Schunk}

For interactive plots, the confidence regions are automatically
detected. When the user clicks on the ROC curve, the confidence region
for the TPF and FPF is overlaid using a grey rectangle. The label and
region stick until the next click.

\begin{Schunk}
\begin{Sinput}
R> cat(
+   export_interactive_roc(ci.rocplot, 
+                          prefix = "aci")
+   )
\end{Sinput}
\end{Schunk}

For use in print, I pass a small vector of cutoff locations at which to
display the confidence regions. This is shown in Figure \ref{conf}.

\begin{Schunk}
\begin{Sinput}
R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
+                  ci.at = c(-.5, .5, 2.1))
\end{Sinput}
\begin{figure}
\includegraphics{figure/printci-1} \caption{Illustration of \pkg{plotROC} plot with exact confidence regions. \label{conf}}\label{fig:printci}
\end{figure}
\end{Schunk}

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
plotted. Therefore, themes and annotations can be added in the usual
\pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme,
title, axis label, and AUC annotation is shown in Figure \ref{annotate}.

\begin{Schunk}
\begin{Sinput}
R> library(ggplot2)
R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
+                  ci.at = c(-.5, .5, 2.1)) + 
+   theme_grey() + 
+   geom_abline(intercept = 0, slope = 1, color = "white") + 
+   ggtitle("Themes and annotations") + 
+   annotate("text", x = .75, y = .25, 
+            label = "AUC = 0.80") +
+   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
\end{Sinput}
\begin{figure}
\includegraphics{figure/print2-1} \caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC} objects. \label{annotate}}\label{fig:print2}
\end{figure}
\end{Schunk}

@@

By default \code{calculate_roc} computes the empirical ROC curve. There
are other estimation methods out there, as I have summarized in the
introduction. Any estimation method can be used, as long as the cutoff,
the TPF and the FPF are returned. Then you can simply pass those values
in a data frame to the \code{ggroc} function. For example, let us use
the binormal method to create a smooth curve. This approach assumes that
the test distribution is normal conditional on disease status.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, 1, .5)
R> M.ex <- rnorm(100, mean = D.ex, sd = .5)
R> 
R> mu1 <- mean(M.ex[D.ex == 1])
R> mu0 <- mean(M.ex[D.ex == 0])
R> s1 <- sd(M.ex[D.ex == 1])
R> s0 <- sd(M.ex[D.ex == 0])
R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
R> 
R> binorm.roc <- data.frame(c = c.ex, 
+                              FPF = pnorm((mu0 - c.ex)/s0), 
+                              TPF = pnorm((mu1 - c.ex)/s1)
+                              )
\end{Sinput}
\end{Schunk}

Then I can pass this \code{data.frame} to the \code{ggroc} function as
before. The example is shown in Figure \ref{binorm}.

\begin{Schunk}
\begin{Sinput}
R> binorm.plot <- ggroc(binorm.roc, label = "Binormal")
R> plot_journal_roc(binorm.plot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
\end{figure}
\end{Schunk}

\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method.
Objects as returned by the function \code{rocr::performance} can be
passed directly to \code{ggroc}:

\begin{Schunk}
\begin{Sinput}
R> library(ROCR)
R> rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
R> rocr.plot <- ggroc(rocr.est, label = "ROCR object")
\end{Sinput}
\end{Schunk}

>>>>

\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. I reviewed a sample of
  ROC curve plots from the major oncology journals in order to assess
  current trends in usage and design elements. My review suggests that ROC
  curve plots are often ineffective as statistical charts and that poor
  design obscures the relevant information the chart is intended to
  display. I describe my new \proglang{R} package that was created to
  address the shortcomings of existing tools. The package has functions to
  create informative ROC curve plots, with sensible defaults and a simple
  interface, for use in print or as an interactive web-based plot. A web
  application was developed to reach a broader audience of scientists who
  do not use \proglang{R}.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves
originated in the field of signal detection theory .

For a continuous measurement that I denote as \(M\), convention dictates
that a test positive is defined as \(M\) equalling or exceeding some
fixed threshold \(c\): \(M \geq c\). The goal of ROC analysis is to
evaluate the classification accuracy of \(M\) in reference to the binary
outcome \(D\), which takes possible values \(0\) (negative) or \(1\)
(positive). It is implicitly assumed that the subpopulation having
\(D = 1\) tends to have larger values of \(M\) compared to the
subpopulation with \(D = 0\). The classification accuracy of \(M\) can
be evaluated by considering the confusion matrix (Table \ref{confus}).
The confusion matrix cross-classifies the predicted outcome \(M \geq c\)
versus the true outcome \(D\). The four cells of the matrix correspond
to the possible classification outcomes: a true positive, a false
positive, a true negative, and a false negative. ROC analysis assesses
the trade-offs between the test's fraction of true positives versus the
false positives.

\begin{table}
\begin{tabular}{l|c|c|c}
\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
\cline{2-3}
\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
\cline{2-3}
 $M \geq c$ & true positive & false positive & test positive\\
\cline{2-3}
 $M < c$ & false negative & true negative & test negative\\
\cline{2-3}
\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
\end{tabular}
\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
\end{table}

@@

\[ TPF(c) = P\{ M \geq c | D = 1 \} \]

@@

\[ FPF(c) = P\{ M \geq c | D = 0 \} \]

Since the cutoff \(c\) is not fixed in advance, one can plot the TPF
against the FPF for all possible values of \(c\). This is exactly what
the ROC curve is, a plot of \(FPF(c)\) on the \(x\) axis and \(TPF(c)\)
along the \(y\) axis as \(c\) varies. A useless test that is not
informative at all in regards to the disease status has
\(TPF(c) = FPF(c)\) for all \(c\). The ROC plot of a useless test is
thus the diagonal line. A perfect test that is completely informative
about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for at least
one value \(c\). If the assumption that support of \(M | D = 1\) is
greater than that of \(M | D = 0\) holds, then the ROC curve will lie in
the upper left quadrant above the diagonal line, however this may not be
the case in a particular sample.

@@

\[ \widehat{TPF(c)} = \frac{\sum_{i = 1}^n 1\{M_i \geq c\} \cdot 1\{D_i = 1\}}{\sum_{i=1}^n 1\{D_i = 1\}}, \]

@@

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
the interesting approach of encoding the cutoff values as a separate
color scale along the ROC curve itself. A legend for the color scale is
placed along the vertical axis on the right of the plotting region. The
\pkg{pROC} package \citep{pROC} provides an option for plotting cutoff
labels (\code{print.thres = TRUE}) and is mainly focused on estimating
confidence intervals and regions for restricted ranges of the ROC curve.
The plotting methods therein use the base \proglang{R} plotting
functions to create nice displays of the curves along with shaded
confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
\citep{ggplot2} plotting library to create clear, informative ROC plots,
with interactive features for use on the web, and sensible defaults for
use in print.

\subsection{Motivation and Literature
Review}\label{motivation-and-literature-review}

@@

I start by creating an example data set. There are 2 markers, one that
is moderately predictive and one that is not as predictive.

\begin{Schunk}
\begin{Sinput}
R> set.seed(2529)
R> D.ex <- rbinom(200, size = 1, prob = .5)
R> M1 <- rnorm(200, mean = D.ex, sd = .65)
R> M2 <- rnorm(200, mean = D.ex, sd = 1.5)
R> 
R> test <- data.frame(D = D.ex, D.str = c("Healthy", "Ill")[D.ex + 1], 
+                    M1 = M1, M2 = M2, stringsAsFactors = FALSE)
\end{Sinput}
\end{Schunk}

\subsubsection{The Roc Geom}\label{the-roc-geom}

As of version 1.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
the previously internal functions \code{Geom} and \code{Stat}. This
enables developers to create their own statistical tranformations and
geometric layers, while enjoying all of the other features of
\pkg{ggplot2}. We have implemented the empirical ROC curve estimate and
the calculation of exact confidence regions as statistical
transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
have also defined geometric layers for the ROC curve and confidence
regions for the ROC curve: \code{geom_roc} and \code{geom_rocci},
respectively. For further discussion and details of the grammar of
graphics as implemented in \pkg{ggplot2}, we refer readers to
\citet{wickham2010layered} and the \pkg{ggplot2} vignettes.

To use the ROC geometric layer, I use the \code{ggplot} function to
define the aesthetic mappings, and the \code{geom_roc} function to add
an ROC curve layer. The \code{geom_roc} function requires the named
aesthetics \code{d} for disease status, and \code{m} for marker. By
default, the ROC geom and stat are linked, so that when \code{geom_roc}
is called, \code{stat_roc} does the computation, and when
\code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
The disease status need not be coded as 0/1, but if it is not,
\code{stat_roc} assumes (with a warning) that the lowest value in sort
order signifies disease-free status.

\begin{Schunk}
\begin{Sinput}
R> basicplot <- ggplot(test, aes(d = D, m = M1)) + geom_roc()
\end{Sinput}
\end{Schunk}

The \code{geom_roc} layer includes the ROC curve line combined with
points and labels to display the values of the biomarker at the
different cutpoints. It accepts the argument \code{n.cuts} to define the
number of cutpoints to display along the curve. Labels can be supressed
by using \code{n.cuts = 0} or \code{labels = FALSE}, however points will
be displayed in the latter case. The size of the labels and the number
of significant digits can be adjusted with \code{labelsize} and
\code{labelround}, respectively.

\begin{Schunk}
\begin{Sinput}
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
\end{Sinput}
\end{Schunk}

\subsubsection{Confidence regions and the Rocci
Geom}\label{confidence-regions-and-the-rocci-geom}

It is common to compute confidence regions for points on the ROC curve
using the \citet{clopper1934use} exact method. Briefly, exact confidence
intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
\citet{pepe2003statistical}, the cross-product of these intervals yields
a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
pair.

This is implemented in the \code{stat_rocci} and displayed as a
\code{geom_rocci} layer. These both require the same aesthetics as the
ROC geom, \code{d} for disease status and \code{m} for marker. By
default, a set of 3 evenly spaced points along the curve are chosed to
display confidence regions. Points corresponding to the confidence
regions are distiguished from the others with a different symbol. You
can select points by passing a vector of values in the range of \code{m}
to the \code{ci.at} argument. By default, the significance level
\(\alpha\) is set to 0.05, this can be changed using the
\code{sig.level} option. An example is shown in Figure \ref{ciex}.

\begin{Schunk}
\begin{Sinput}
R> basicplot + geom_rocci()
\end{Sinput}
\begin{figure}
\includegraphics{figure/test-a-ci-1} \caption{Illustration of \pkg{plotROC} with exact confidence regions. \label{ciex}}\label{fig:test-a-ci}
\end{figure}
\begin{Sinput}
R> ## 
R> ## basicplot + geom_rocci(sig.level = .01)
R> ## ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0) +
R> ##   geom_rocci(ci.at = quantile(M1, c(.1, .4, .5, .6, .9)))
\end{Sinput}
\end{Schunk}

\subsubsection{Styles and Labels}\label{styles-and-labels}

The same objects like \code{basicplot} with Roc and/or Rocci layers can
be treated like any other ggplot objects. It can be printed to display
the figure, and other layers can be added to the plot. We provide the
function \texttt{style\_roc()} which is a layer containing a theme,
modified gridlines, and axes. Adding the \texttt{style\_roc()} layer to
the ggplot object creates a plot with sensible defaults for use in
print. This function has options for the number and location of major
and minor breaks, addition of the diagnoal guideline, axis labels, and
any theme object created for use with ggplot2 can be supplied.

The \code{direct_label} function takes a ggplot object as an argument
and annotates the figure with a direct label with a automatically chosen
location. It attempts to intellegently select an appropriate location
for the label, but the location can be adjusted with
\code{nudge_x, nudge_y} and \code{label.angle}. If the \code{labels}
argument is NULL, it will take the name from the mapped aesthetic. A
simple example with the default options is shown in Figure \ref{first}.

\begin{Schunk}
\begin{Sinput}
R> direct_label(basicplot) + style_roc()
\end{Sinput}
\begin{figure}
\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
\end{figure}
\begin{Sinput}
R> ## 
R> ## basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
\end{Sinput}
\end{Schunk}

\subsubsection{Interactive ROC plots}\label{interactive-roc-plots}

The \code{basicplot} object, which is of class ggplot, can be used to
create an interactive plot and display it in the Rstudio viewer or
default web browser by passing it to the \code{plot_interactive_roc}
function. Give the function an optional path to an html file as an
argument called \code{file} to save the interactive plot as a complete
web page. A screen shot of an interactive plot is shown in Figure
\ref{interact}. Hovering over the display shows the cutoff value at the
point nearest to the cursor. Clicking makes the cutoff label stick until
the next click, and if confidence regions are available, clicks will
also display those as grey rectangles. By default,
\code{plot_interactive_roc} removes any existing Rocci geom and adds a
high-density layer of confidence regions. This can be suppressed by
using the \code{add.cis = FALSE} option. The points and labels layer of
the Roc geom can be hidden by using the \code{hide.points = TRUE}
option. Then, points and labels will be displayed only when the mouse is
hovering over the plotting region. Also by default, the \code{style_roc}
function is applied, the settings can be modified by passing a call to
that function.

\begin{Schunk}
\begin{Sinput}
R> plot_interactive_roc(basicplot)
R> plot_interactive_roc(basicplot, hide.points = TRUE)
R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
\end{Sinput}
\end{Schunk}

@@

Hovering over the display shows the cutoff value at the point nearest to
the cursor. Clicking makes the cutoff label stick until the next click,
and if confidence regions are available, clicks will also display those
as grey rectangles. The confidence regions are automatically detected.
When the user clicks on the ROC curve, the confidence region for the TPF
and FPF is overlaid using a grey rectangle. The label and region stick
until the next click.

An interactive ROC plot can be exported by using the
\code{export_interactive_roc} function, which returns a character string
containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
\proglang{JavaScript} source can be omitted by using the
\code{omit.js = TRUE} option. You may wish to do this when there are
multiple interactive figures in a single document; the source only needs
to be included once. The character string can be copy-pasted into an
html document, or better yet, incorporated directly into a dynamic
document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
is necessary to use the \code{cat} function on the results and use the
chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so
that the interactive plot is displayed correctly. For documents that
contain multiple interactive plots, it is necessary to assign each plot
a unique name using the \code{prefix} argument of
\code{export_interactive_roc}. This is necessary to ensure that the
\proglang{JavaScript} code manipulates the correct svg elements. For
examples of interactive plots and how to incorporate them into
\pkg{knitr} documents, see the package vignette
(\code{vignette("examples", package = "plotROC")}) or the web page
https://sachsmc.github.io/plotROC/. The next code block shows an example
\pkg{knitr} chunk that can be used in an .Rmd document to display an
interactive plot.

\begin{Code}
```{r int-no, results = 'asis', fig.keep = 'none'}
cat(
  export_interactive_roc(basicplot, 
                        prefix = "a")
  )
```
\end{Code}

@@

If you have grouping factors in your dataset, or you have multiple
markers measured on the same subjects, you may wish to plot multiple ROC
curves on the same plot. \pkg{plotROC} fully supports faceting and
grouping done by \pkg{ggplot2}. In out example dataset, we have 2
markers measured in a paired manner:

\begin{Schunk}
\begin{Sinput}
R> head(test)
\end{Sinput}
\begin{Soutput}
  D   D.str         M1          M2
1 1     Ill 1.48117155 -2.50636605
2 1     Ill 0.61994478  1.46861033
3 0 Healthy 0.57613345  0.07532573
4 1     Ill 0.85433197  2.41997703
5 0 Healthy 0.05258342  0.01863718
6 1     Ill 0.66703989  0.24732453
\end{Soutput}
\end{Schunk}

These data are in wide format, with the 2 markers going across 2
columns. \pkg{ggplot2} requires long format, with the marker result in a
single column, and a third variable identifying the marker. We provide
the convenience function \code{melt_roc} to perform this transformation.
The arguments are the data frame, a name or index identifying the
disease status column, and a vector of names or indices identifying the
the markers. Optionally, the names argument gives a vector of names to
assign to the marker, replacing their column names. The result is a data
frame in long format.

\begin{Schunk}
\begin{Sinput}
R> longtest <- melt_roc(test, "D", c("M1", "M2"))
R> head(longtest)
\end{Sinput}
\begin{Soutput}
    D          M name
M11 1 1.48117155   M1
M12 1 0.61994478   M1
M13 0 0.57613345   M1
M14 1 0.85433197   M1
M15 0 0.05258342   M1
M16 1 0.66703989   M1
\end{Soutput}
\end{Schunk}

Then, the dataset can be passed to the \code{ggplot} function, with the
marker name given as a grouping or faceting variable.

\begin{Schunk}
\begin{Sinput}
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
R> ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
R> pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
+   geom_roc(show.legend = FALSE) + style_roc()
R> direct_label(pairplot)
R> 
R> pairplot + geom_rocci()
R> pairplot + geom_rocci(linetype = 1)
\end{Sinput}
\end{Schunk}

Showing multiple curves is also useful when there is a factor that
affects the classification accuracy of the test. Let's create another
example dataset.

\begin{Schunk}
\begin{Sinput}
R> D.cov <- rbinom(400, 1, .5)
R> gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
R> M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
R> 
R> test.cov <- data.frame(D = D.cov, gender = gender, M = M.diff)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
R> bygend <- ggplot(test.cov, aes(d = D, m = M, color = gender)) + 
+   geom_roc(show.legend = FALSE)
R> direct_label(bygend) + style_roc()
\end{Sinput}
\begin{figure}
\includegraphics{figure/covplot-1} \caption{Illustration of \pkg{plotROC} with multiple curves. \label{multi}}\label{fig:covplot}
\end{figure}
\end{Schunk}

Interactive versions of the plots with grouping and faceting are fully
supported.

@@

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
plotted. Therefore, themes and annotations can be added in the usual
\pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme,
title, axis label, and AUC annotation is shown in Figure \ref{annotate}.
\pkg{plotROC} provides the convenience function \code{calc_auc} that
takes a ggplot object that has an Roc layer, extracts the data, and
calculates the AUC.

\begin{Schunk}
\begin{Sinput}
R> annotate <- basicplot + style_roc(theme = theme_grey) +
+   theme(axis.text = element_text(colour = "blue")) +
+   ggtitle("Themes and annotations") + 
+   annotate("text", x = .75, y = .25, 
+            label = paste("AUC =", round(calc_auc(basicplot)["AUC"], 2))) +
+   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{figure/print2-1} \caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC} objects. \label{annotate}}\label{fig:print2}
\end{figure}
\end{Schunk}

The results are compatible with all other \pkg{ggplot2} layers and
functions, and interactive versions are supported as long as there is an
Roc layer present.

@@

By default \code{calculate_roc} computes the empirical ROC curve. There
are other estimation methods out there, as I have summarized in the
introduction. Any estimation method can be used, as long as the cutoff,
the TPF and the FPF are returned. Then you can simply pass those values
in a data frame to the \code{ggplot} function and override the default
statistical tranformation to \code{stat_identity}. For example, let us
use the binormal method to create a smooth curve. This approach assumes
that the test distribution is normal conditional on disease status.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- test$D
R> M.ex <- test$M1
R> mu1 <- mean(M.ex[D.ex == 1])
R> mu0 <- mean(M.ex[D.ex == 0])
R> s1 <- sd(M.ex[D.ex == 1])
R> s0 <- sd(M.ex[D.ex == 0])
R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
R> 
R> binorm.roc <- data.frame(c = c.ex, 
+                              FPF = pnorm((mu0 - c.ex)/s0), 
+                              TPF = pnorm((mu1 - c.ex)/s1)
+                              )
R> 
R> binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
+   geom_roc(stat = "identity") + style_roc(theme = theme_grey)
\end{Sinput}
\end{Schunk}

The example is shown in Figure \ref{binorm}. Interactive plots with
\code{stat = 'identity'} are not currently supported.

\begin{Schunk}
\begin{Sinput}
R> binorm.plot
\end{Sinput}
\begin{figure}
\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
\end{figure}
\end{Schunk}
